Reference Tracking Strategies of Deaf Adult Signers in Turkish Sign Language

### 2023-05-01 revised_manuscript.tex
updated

However, this particular observation is not peculiar to spoken
languages. There is an emerging body of literature that reports the
distribution of REs in signed narratives (Ahlgren and Bergman, 1994;
Czubek, 2017; Ferrara et al., 2022; Frederiksen and Mayberry, 2016;
Hodge et al., 2019; Morgan, 2005; Perniss and Özyürek, 2015; Swabey,
2002). Building on the work by Gundel et al. (1993), Swabey (2002)
demonstrated that, like other languages such as Russian, Chinese, and
Japanese, American Sign Language (ASL) signers use bare nouns for
introduced and re-introduced discourse contexts whereas zero anaphora is
used for referents which are ``in-focus'' or very accessible to the
addressee. Overt REs like pronouns or nouns are disprefered for such
referents. Swabey (2002) concluded that ``native'' signers in her study
abided by the Gricean maxim of quantity and thus used more economic
forms of reference tracking. Similar findings have been reported for
German Sign Language (Deutsche Gebärdensprache - DGS) (Perniss and
Özyürek, 2015) and British Sign Language (BSL) (Morgan, 2005). It is
important to note, however, that this research either collapsed DoD and
DoH signers into a single group and labeled them as ``native-like'' or
explicitly compared native deaf signers with late deaf signers (with
delayed exposure to a sign language) in terms of narrative competence.
Most recently, Ferrara et al. (2022) examined corpora including
video-recorded narratives from five different sign languages including
Irish Sign Language, Finnish Sign Language, Norwegian Sign Language and
Swedish Sign Language and investigated how these languages established
and maintained referents. Like other observations, this research showed
that, across each of these languages, bare nouns were used to introduce
new referents and zero anaphora or more reduced forms were preferred for
referent maintenance.

@@

In TİD and many other sign languages, it is possible to use bare nouns
as in \ref{ex:2a} as well as fingerspelled nouns as in \ref{ex:2b}.
Fingerspelling refers to the use of the manual aphabet to spell all or
part of the letters of a word borrowed from a surrounding written
language (Kubuş, 2008). Nouns can also be modified by a prenominal or
postnominal extended index finger pointing sign, glossed as IX,
functioning as a determiner or a demonstrative as in \ref{ex:2c}
(Maclaughlin, 1997) or by a modifying classifier within the same NP as
in \ref{ex:2d} (Frederiksen and Mayberry,
2016).\footnote{Spoken languages use pronouns, determines and demonstratives where sign languages use a pointing sign, usually glossed as IX, which assumes similar functions (Lillo-Martin and Klima, 1990). In addition, the pointing sign is used to referentially anchor a referent to a location in signing space.}

@@

We tested 29 deaf adult signers, 15 DoD signers with exposure to TİD
from birth from deaf caregivers (e.g., parents or siblings), 14 DoH
signers who had hearing caregivers and were exposed to TİD between ages
3 and 17. (\emph{M}\textsubscript{Age of acquisition} = 7.7 years,
\emph{SD} = 3.3 years). As a result, the criterion to be classified as a
DoH signer was to have hearing and non-signing caregivers. All the
signers satisfying this criterion were regarded as
such.\footnote{Although we did not set a cut-off point for nativeness, all the deaf of hearing participants indicated that they learned TİD after age 3. In order to capture in-group variation among DoH signers, we run an additional supplementary analysis in which we observed very wide credible intervals due to the small group size, rendering the results uninformative. For the analysis, see \ref{section:d}.}
All participants indicated that their preferred language of
communication was TİD which they used for most contexts. Detailed
information about the linguistic and educational background of the
participants can be found in \ref{section:a}. 1 DoD signer was excluded
from the data analysis because they self-reported to have signing deaf
parents but also only used home sign (a more crude form of
communication) at home. As a result, we present data from only 14 DoD
signers.

@@

Following Berman and Slobin (2013), we first identified the clause
boundaries in each sentence indicated by the presence of predicates and
informative non-manual markers (e.g., head nod). On the second tier, we
coded the discourse status for each referent. Having adapted the
classification of Gullberg (2006), we considered all referents mentioned
for the first time as introduced regardless of their syntactic function
in the sentence. Only subjects were coded for maintenance and
re-introduction. We coded subject referents as maintained only if they
had been referred to in the previous clause. To this end, the referents
which had been discontinued across one or more sentences were all
regarded re-introduced into the discourse. On a third tier, we had the
RE type for a single reference with the main categories nominal,
pronominal, zero anaphora, and classifier. Under nominal, we identified
bare nouns, modified nouns, and fingerspelled nouns. For classifiers, we
distinguished among three types of handshapes: WECL, BPCL, and ExtCL.
Finally, we had CA, agreement, plain tags for zero anaphora. We were
unable to observe any instance of HCLs occurring in isolation (i.e., not
part of CA). We should also highlight here that we coded referent status
as nominal if the clause included an explicit nominal referent. For
example, if constructed action, an index sign, or a classifier preceded
or followed its explicitly signed nominal referent, the clause was
regarded as having nominal reference (i.e., modified noun). The
annotation tiers are presented in Table \ref{tab:tab-2}.

@@

\begin{figure}
\centering
\includegraphics{revised_manuscript_files/figure-latex/fig:count-plot-1.pdf}
\caption{Referent counts by Discourse Status (Introduction, Maintanence,
Re-introduction), RE Type (Nominal, Pronominal, Classifier, Zero
Anaphora), and Participant Group (DoD, DoH)}
\end{figure}

@@

As for the distribution of the zero anaphora, approximately 90\% of the
plain verbs were used to maintain a previous referent and this makes
these constructions least implicit REs used for TİD\footnote{As in other
  sign languages, in TİD, the lexical meaning of signs has an effect on
  the morphological grouping. For instance, usually (di)transitive
  agreement verbs denote some sort of physical or metaphorical transfer
  of an object or abstract notion (Meir, 1999). As for plain verbs, both
  phonology of a sign and its semantics have an effect on morphology.
  For instance, \textsc{eat} in TİD is signed nearby the mouth and since
  it needs to remain in that phonological location, it is not suitable
  for agreement and thus it is regarded as a plain verb. Likewise the
  experiencer verb \textsc{love} is signed on the torso and as such it
  does not take part in agreement. These generalizations noted we should
  also be cautious in labelling a verb as categorially one type or the
  other. For instance, some intransitive plain verbs like \textsc{fall}
  or \textsc{jump} in TİD may display spatial person agreement with the
  single argument in the clause (Gökgöz and Sevgi, 2020). See Lourenço
  and Wilbur (2018) also for a similar discussion in Brazilian Sign
  Language.}. Agreement verbs and CA were the following implicit markers
of reference, 0.77 and 0.72 of which were used for maintained contexts,
respectively. This contrasts with some of the research (Czubek, 2017) in
which agreement verbs ranked lowest on the scale, indicating that these
constructions were more implicit compared to plain verbs. However, there
seems to be a common ground for the position of CA markers which have
been claimed to rank higher than other markers of zero anaphora as in
the present study (Barberà and Massó, 2009; Czubek, 2017; Frederiksen
and Mayberry, 2016). In Figure \ref{fig:scale}, we report the
distribution of REs in TİD as a function of referent accessibility and
compare our accessibility hierarchy with that of (Frederiksen and
Mayberry, 2016) for ASL. In the hierarchies, the leftmost REs (e.g.,
modified and bare nouns for TİD) are used when the accessibility is low.
If the referent is already active in the addressee's mind and its
accessibility is high, then the rightmost markers (e.g., agreement and
plain verbs for TİD) are selected.

\begin{figure}
\caption{Comparing the present study and Frederiksen and Mayberry (2016) in terms of proposed accessibility hierarchies}
\label{fig:scale}
\tikzset{every picture/.style={line width=0.75pt}} %set default line width to 0.75pt

@@

Both hierarchies illustrated in Figure \ref{fig:scale} treat nominals
and ExtCLs as markers of lowly accessibly referents. Likewise, in both
hierarchies, as markers of highly accessible referents, different types
of zero anaphora markers are further distinguished. Plain verbs rank the
lowest on the scales, preceded by agreement verbs and CA. On a similar
ground, both scales lack pronouns, which occurred very infrequently in
the data. Although Frederiksen and Mayberry (2016) did not include other
classifier types in the hierarchy, we propose that WECLs and BPCLs must
rank higher than zero anaphora but lower than ExtCLs that were mainly
used by interlocutors to introduce a referent into the discourse
context.

@@

While DoH signers were only partially over-redundant in maintained
contexts, they were more so in introduced contexts. In line with our
study, a tendency for limited over-redundancy has been suggested for
late signing children in DGS (Becker, 2009) and late signing adults in
BSL specifically for the use of CA (Cormier et al., 2013). As for TİD,
to our knowledge, there is only one study (Gür, 2018) that directly
aimed at examining age of acquisition effects on adult signers'
narrative skills. Gür (2018) observed that late signing deaf adults and
children set the scene less frequently and used CA less accurately
compared to their native counterparts. However, this particular
investigation did not present any data on the distribution of REs
between two acquisition groups. In the present study, the finding that
DoD signers preferred to use null arguments even for the first mention
of a referent (19 in total) is surprising given some of the previous
literature in the topicality and saliency framework (Cormier et al.,
2013; Swabey, 2002). These research reported very little or no use of
zero anaphora by DoD signers for introduced contexts.

@@

@@

@@

@@

>>>>

However, this particular observation is not peculiar to spoken
languages. There is an emerging body of literature that reports the
distribution of REs in signed narratives (Ahlgren and Bergman, 1994;
Czubek, 2017; Ferrara et al., 2022; Frederiksen and Mayberry, 2016;
Garcia et al., 2018; Garcia and Sallandre, 2014; Hodge et al., 2019;
Morgan, 2005; Perniss and Özyürek, 2015; Swabey, 2002). Building on the
work by Gundel et al. (1993), Swabey (2002) demonstrated that, like
other languages such as Russian, Chinese, and Japanese, American Sign
Language (ASL) signers use bare nouns for introduced and re-introduced
discourse contexts whereas zero anaphora is used for referents which are
``in-focus'' or very accessible to the addressee. Overt REs like
pronouns or nouns are disprefered for such referents. Swabey (2002)
concluded that ``native'' signers in her study abided by the Gricean
maxim of quantity and thus used more economic forms of reference
tracking. Similar findings have been reported for German Sign Language
(Deutsche Gebärdensprache - DGS) (Perniss and Özyürek, 2015) and British
Sign Language (BSL) (Morgan, 2005). It is important to note, however,
that this research either collapsed DoD and DoH signers into a single
group and labeled them as ``native-like'' or explicitly compared native
deaf signers with late deaf signers (with delayed exposure to a sign
language) in terms of narrative competence. Most recently, Ferrara et
al. (2022) examined corpora including video-recorded narratives from
five different sign languages including Irish Sign Language, Finnish
Sign Language, Norwegian Sign Language and Swedish Sign Language and
investigated how these languages established and maintained referents.
Like other observations, this research showed that, across each of these
languages, bare nouns were used to introduce new referents and zero
anaphora or more reduced forms were preferred for referent maintenance.

@@

In TİD and many other sign languages, it is possible to use bare nouns
as in \ref{ex:2a} as well as fingerspelled nouns as in \ref{ex:2b}.
Fingerspelling refers to the use of the manual alphabet to spell all or
part of the letters of a word borrowed from a surrounding written
language (Kubuş, 2008). Nouns can also be modified by a prenominal or
postnominal extended index finger pointing sign, glossed as IX,
functioning as a determiner or a demonstrative as in \ref{ex:2c}
(Maclaughlin, 1997) or by a modifying classifier within the same NP as
in \ref{ex:2d} (Frederiksen and Mayberry,
2016).\footnote{Spoken languages use pronouns, determines and demonstratives where sign languages use a pointing sign, usually glossed as IX, which assumes similar functions (Lillo-Martin and Klima, 1990). In addition, the pointing sign is used to referentially anchor a referent to a location in signing space.}

@@

We tested 29 deaf adult signers, 15 DoD signers with exposure to TİD
from birth from deaf caregivers (e.g., parents or siblings), 14 DoH
signers who had hearing caregivers and were exposed to TİD between ages
3 and 17. (\emph{M}\textsubscript{Age of acquisition} = 7.7 years,
\emph{SD} = 3.3 years). As a result, the criterion to be classified as a
DoH signer was to have hearing and non-signing caregivers. All the
signers satisfying this criterion were regarded as
such.\footnote{Although we did not set a cut-off point for nativeness, all the deaf of hearing participants indicated that they learned TİD after age 3. In order to capture in-group variation among DoH signers, we run an additional supplementary analysis in which we observed very wide credible intervals due to the small group size, rendering the results uninformative. For the analysis, see \ref{section:d}.}
Detailed information about the linguistic and educational background of
the participants can be found in \ref{section:a}.

All participants indicated that their preferred language of
communication was TİD which they used for most contexts. 1 DoD signer
was excluded from the data analysis because they self-reported to have
signing deaf parents but also only used home sign (a more crude form of
communication) at home. As a result, we present data from only 14 DoD
signers.

@@

Following Berman and Slobin (2013), we first identified the clause
boundaries in each sentence indicated by the presence of predicates and
informative non-manual markers (e.g., head nod). On the second tier, we
coded the discourse status for each referent. Having adapted the
classification of Gullberg (2006), we considered all referents mentioned
for the first time as introduced regardless of their syntactic function
in the sentence. Only subjects were coded for maintenance and
re-introduction. We coded subject referents as maintained only if they
had been referred to in the previous clause. To this end, the referents
which had been discontinued across one or more sentences were all
regarded re-introduced into the discourse. On a third tier, we had the
RE type for a single reference with the main categories nominal,
pronominal, zero anaphora, and classifier. Under nominal, we identified
bare nouns, modified nouns, and fingerspelled nouns. For classifiers, we
distinguished among three types of handshapes: WECL, BPCL, and ExtCL.
Finally, we had CA\footnote{Note that CA can also be further subdivided
  into overt, reduced, and subtle CA, following (Cormier et al., 2015).
  In overt CA, the signer is ``fully in character'' and does not use any
  other lexical or classifier constructions. In reduced CA, the signer
  can involve a small amount of narration or lexical material while
  tracking a referent using CA. However, subtle CA refers to cases where
  the referent is primarily tracked by a non-CA construction (e.g.,
  nominal, pronominal, verbal or classifier constructions) but the
  signer subtly uses a few CA articulators at the same time (e.g.,
  enacting the eye-gaze of the character). We only coded overt and
  reduced types as reference tracking by CA, and disregarded the subtle
  type. We'd also like to acknowledge that mixed CA, or simultaneous
  tracking of two or multiple references through CA and other tools can
  be possible (Dudis, 2004; Liddell, 2003). In such cases, as we explain
  here, we only coded only the subject of a clause for maintenance and
  re-introduction. If the referent was in subject position and tracked
  by using CA, we considered this an instance of CA, as well.},
agreement, plain tags for zero anaphora. We were unable to observe any
instance of HCLs occurring in isolation (i.e., not part of CA). We
should also highlight here that we coded referent status as nominal if
the clause included an explicit nominal referent. For example, if
constructed action, an index sign, or a classifier preceded or followed
its explicitly signed nominal referent, the clause was regarded as
having nominal reference (i.e., modified noun). The annotation tiers are
presented in Table \ref{tab:tab-2}.

@@

\begin{figure}
\centering
\includegraphics{revised_manuscript_files/figure-latex/fig-count-plot-1.pdf}
\caption{Referent counts by Discourse Status (Introduction, Maintanence,
Re-introduction), RE Type (Nominal, Pronominal, Classifier, Zero
Anaphora), and Participant Group (DoD, DoH)}
\end{figure}

@@

As for the distribution of the zero anaphora, approximately 90\% of the
plain verbs were used to maintain a previous referent and this makes
these constructions least implicit REs used for TİD\footnote{As in other
  sign languages, in TİD, the lexical meaning of signs has an effect on
  the morphological grouping. For instance, usually (di)transitive
  agreement verbs denote some sort of physical or metaphorical transfer
  of an object or abstract notion (Meir, 1999). As for plain verbs, both
  phonology of a sign and its semantics have an effect on morphology.
  For instance, \textsc{eat} in TİD is signed nearby the mouth and since
  it needs to remain in that phonological location, it is not suitable
  for agreement and thus it is regarded as a plain verb. Likewise the
  experiencer verb \textsc{love} is signed on the torso and as such it
  does not take part in agreement. These generalizations noted we should
  also be cautious in labelling a verb as categorially one type or the
  other. For instance, some intransitive plain verbs like \textsc{fall}
  or \textsc{jump} in TİD may display spatial person agreement with the
  single argument in the clause (Gökgöz and Sevgi, 2020). See Lourenço
  and Wilbur (2018) also for a similar discussion in Brazilian Sign
  Language.}. Agreement verbs and CA were the following implicit markers
of reference, 0.77 and 0.72 of which were used for maintained contexts,
respectively. This contrasts with some of the research (Czubek, 2017) in
which agreement verbs ranked lowest on the scale, indicating that these
constructions were more implicit compared to plain verbs. However, there
seems to be a common ground for the position of CA markers which have
been claimed to rank higher than other markers of zero anaphora as in
the present study (Barberà and Massó, 2009; Czubek, 2017; Frederiksen
and Mayberry, 2016). In Figure \ref{fig-scale}, we report the
distribution of REs in TİD as a function of referent accessibility and
compare our accessibility hierarchy with that of (Frederiksen and
Mayberry, 2016) for ASL. In the hierarchies, the leftmost REs (e.g.,
modified and bare nouns for TİD) are used when the accessibility is low.
If the referent is already active in the addressee's mind and its
accessibility is high, then the rightmost markers (e.g., agreement and
plain verbs for TİD) are selected.

\begin{figure}
\caption{Comparing the present study and Frederiksen and Mayberry (2016) in terms of proposed accessibility hierarchies}
\label{fig-scale}
\tikzset{every picture/.style={line width=0.75pt}} %set default line width to 0.75pt

@@

Both hierarchies illustrated in Figure \ref{fig-scale} treat nominals
and ExtCLs as markers of lowly accessibly referents. Likewise, in both
hierarchies, as markers of highly accessible referents, different types
of zero anaphora markers are further distinguished. Plain verbs rank the
lowest on the scales, preceded by agreement verbs and CA. On a similar
ground, both scales lack pronouns, which occurred very infrequently in
the data. Although Frederiksen and Mayberry (2016) did not include other
classifier types in the hierarchy, we propose that WECLs and BPCLs must
rank higher than zero anaphora but lower than ExtCLs that were mainly
used by interlocutors to introduce a referent into the discourse
context.

@@

While DoH signers were only partially over-redundant in maintained
contexts, they were more so in introduced contexts. In line with our
study, a tendency for limited over-redundancy has been suggested for
late signing children in DGS (Becker, 2009) and late signing adults in
BSL specifically for the use of CA (Cormier et al., 2013). As for TİD,
to our knowledge, there is only one study (Gür, 2018) that directly
aimed at examining age of acquisition effects on adult signers'
narrative skills. Gür (2018) observed that late signing deaf adults and
children set the scene less frequently and used CA less accurately
compared to their native counterparts. However, this particular
investigation did not present any data on the distribution of REs
between two acquisition groups. In the present study, the finding that
DoD signers preferred to use null arguments even for the first mention
of a referent (19 in total) is surprising given some of the previous
literature in the topicality and saliency framework (Cormier et al.,
2013; Ferrara et al., 2022; Swabey, 2002). These research reported very
little or no use of zero anaphora by DoD signers for introduced
contexts.

@@

\leavevmode\vadjust pre{\hypertarget{ref-cormier_rethinking_2015}{}}%
Cormier, Kearsy, Smith, Sandra, Sevcikova-Sehyr, Zed, 2015. Rethinking
constructed action. {SL}\&L 18, 167--204.
doi:\href{https://doi.org/10.1075/sll.18.2.01cor}{10.1075/sll.18.2.01cor}

@@

\leavevmode\vadjust pre{\hypertarget{ref-dudis2004}{}}%
Dudis, Paul G., 2004.
\href{https://www.degruyter.com/view/j/cogl.2004.15.issue-2/cogl.2004.009/cogl.2004.009.xml}{Body
partitioning and real-space blends}. Cognitive Linguistics 15.

@@

\leavevmode\vadjust pre{\hypertarget{ref-garcia_reference_2014}{}}%
Garcia, Brigitte, Sallandre, Marie-Anne, 2014. Reference resolution in
french sign language ({LSF}), in: Cabredo Hofherr, Patricia,
Zribi-Hertz, Anne (Eds.), Crosslinguistic Studies on Noun Phrase
Structure and Reference. Brill, Leiden, pp. 316--364.
doi:\href{https://doi.org/10.1163/9789004261440_012}{10.1163/9789004261440\_012}

\leavevmode\vadjust pre{\hypertarget{ref-garcia2018impersonal}{}}%
Garcia, Brigitte, Sallandre, Marie-Anne, L'Huillier, Marie-Thérèse,
2018. Impersonal human reference in french sign language (LSF). Sign
Language \& Linguistics 21, 307--333.

@@

\leavevmode\vadjust pre{\hypertarget{ref-liddell2003}{}}%
Liddell, Scott K., 2003. Grammar, gesture, and meaning in American Sign
Language. Cambridge University Press.
